---
title: "6 Visualise Uncertainty"
subtitle: Hands-On Exercise for Week 4
author: "KB"
editor: visual
execute:
  freeze: auto 
  warning: false
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    code-line-numbers: true
    toc-depth: 4
---

Published: 4-May-2023

## 6.1 Learning Outcome

A point estimate, such as mean, is a single numerical value that is used to estimate an unknown population parameter. On the other hand, a range estimate is a range of values that is used to estimate an unknown population parameter. A range estimate is useful when a single point estimate is not precise enough, or when we want to communicate the level of uncertainty surrounding the point estimate.

In this exercise, we will explore approaches to visualise the uncertainty of point estimates.

## 6.2 Getting Started

### 6.2.1 Install and load the required r libraries

Install and load the the required R packages. The name and function of the new package that will be used for this exercise is as follow:

-   **ggdist**: provides a range of functions for creating visualizations of probability distributions
-   **knitr**: create high-quality, fully reproducible documents that integrate code, text, and graphics

```{r}
pacman::p_load(tidyverse, plotly, crosstalk, DT, gganimate, ggdist, knitr)
```

### 6.2.2 Import the data

We will be using the same exam scores data-set that was featured in my [Hands-On Exercise for Week 1](https://boomie-boom-isss608.netlify.app/hands-on_ex/hands-on_ex01/hands-on_ex1#import-the-data "Additional info on the data-set").

```{r}
exam <- read_csv('data/Exam_data.csv', show_col_types = FALSE )
```

## 6.3 Visualise the uncertainty of point estimates with ggplot2 package

First, we perform with following on the exam scores data-set:

-   group the observations by RACE,

-   compute the count of observations, mean, standard deviation and standard error of Maths scores by RACE, and

-   assign the output as a tibble data table called \`my_sum

-   print the tibble data using `kable()` function from **knitr**

```{r}
my_sum <- exam %>%
  group_by(RACE) %>%
  summarise(
    n=n(),
    mean=mean(MATHS),
    sd=sd(MATHS)
    ) %>%
  mutate(se=sd/sqrt(n-1))

kable(head(my_sum), format = 'html') 
```

We can use the `geom_errorbar()` function to reveal the standard error of mean maths score by race in the following chart.

```{r}
ggplot(my_sum) +
  geom_errorbar(
    aes(x=RACE, 
        ymin=mean-se, 
        ymax=mean+se), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    linewidth=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  ggtitle("Standard Error of the Mean Math Scores by Race")
```

::: callout-note
## We plotted this before\...

Check out [this plot](https://cosmic-boon.netlify.app/hands-on_ex/hands-on_ex03/hands-on_ex3a#the-onclick-argument) which we did in an earlier exercise.
:::

We can also use the `geom_errorbar()` function to reveal the 95% Confidence Interval of the mean maths score by race in the following chart.

```{r}
#| code-fold: false

# Calculate the lower and upper bound for confidence intervals
my_sum2 <- my_sum %>% 
  mutate(lower_ci = mean - qt(0.975, n-1)*se,
         upper_ci = mean + qt(0.975, n-1)*se)

# Create point estimate plot with error bars
ggplot(my_sum2) +
  geom_errorbar(
    aes(x=reorder(RACE,-mean), 
        ymin=lower_ci, 
        ymax=upper_ci), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    linewidth=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  labs(title="95% Confidence Interval of the Mean Math Scores by Race",
       x = "Race",
       y = "Math Score")
```

### 6.3.1 Visualise the uncertainty of point estimates with interactive error bars

The following interactive error bar plot shows the 99% Confidence Interval of mean maths score by race.

( 🖱️Click on the average score (in red) for the cross-filtering to work)

```{r}
# Prepare the summary table with the relevant stats
my_sum3 <- exam %>%
  group_by(RACE) %>%
  summarise(
    'No. of pupils'=n(),
    'Avg Scores'=mean(MATHS),
    'Std Dev'=sd(MATHS),
    'Std Error' = sd(MATHS)/sqrt(n()-1)
    ) 

# Use highlight_key() to add a unique key to the data frame my_sum3 so that it can be linked to interactive plots later
d <- highlight_key(my_sum3)

# Prepare the error bar plot
p <- ggplot(d) +
  geom_errorbar(
    aes(x=reorder(RACE,-`Avg Scores`), 
        ymin=`Avg Scores` - qt(0.995, `No. of pupils`-1)*`Std Error`, 
        ymax=`Avg Scores` + qt(0.995, `No. of pupils`-1)*`Std Error`), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    linewidth=0.5) +
  geom_point(aes
           (x=RACE, 
            y=`Avg Scores`), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  labs(title="99% Confidence Interval of the Mean Math Scores by Race",
       x = "Race",
       y = "Math Score")

# Convert ggplot to an interactive plotly plot using the ggplotly(), "plotly click" specifies that highlight should be based on click
gg <- highlight(ggplotly(p),on =        
                "plotly_click")

# Create a Bootstrap grid of two columns to house the 2 plots in the ratio of 5:7
bscols(gg,datatable(d,options = list(dom='t')),widths = c(5,7))

```

## 6.4 Visualising Uncertainty using ggdist package

[**ggdist**](https://mjskay.github.io/ggdist/) is an R package that provides a flexible set of ggplot2 geoms and stats designed especially for visualising distributions and uncertainty. It is designed for both frequentist and Bayesian uncertainty visualization, taking the view that uncertainty visualization can be unified through the perspective of distribution visualization:

-   for frequentist models, one visualises confidence distributions or bootstrap distributions (see vignette("freq-uncertainty-vis"));

-   for Bayesian models, one visualises probability distributions (see the tidybayes package, which builds on top of ggdist).

We use the [`stat_pointinterval()`](https://mjskay.github.io/ggdist/reference/stat_pointinterval.html) of **ggdist** to build a visual for displaying distribution of maths scores by race.

```{r}
exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_pointinterval() +   #<<
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Mean Point + Multiple-interval plot") +
  theme_minimal()

```

This [`stat_pointinterval()`](https://mjskay.github.io/ggdist/reference/stat_pointinterval.html) function comes with many arguments. The following plot creates a plot that displays the median math score for each race group, along with an interval estimate for each group. This can help visualize the differences in math scores across different racial groups and also the variability within each group.

```{r}
exam %>%
  ggplot(aes(x = RACE, y = MATHS)) +
  stat_pointinterval(.width = 0.95,
  point_interval = "median_qi"
  ) +
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Mean Point + Multiple-interval plot")
```

We can use the [`stat_gradientinterval()`](https://mjskay.github.io/ggdist/reference/stat_gradientinterval.html) function of **ggdist** to build a visual for displaying distribution of maths scores by race.

```{r}
exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_gradientinterval(   
    fill = "skyblue",      
    show.legend = TRUE     
  ) +                        
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Gradient + interval plot")
```

## 6.5 Visualise Uncertainty with Hypothetical Outcome Plots (HOPs) **\[Can't run\]**

Step 1: Install the **ungeviz** package

```{r}
#| eval: false

devtools::install_github("wilkelab/ungeviz")
```

Step 2: Launch the application in R

```{r}
#| eval: false
library(ungeviz)
```

```{r}
#| eval: false
ggplot(data = exam, 
       (aes(x = factor(RACE), 
            y = MATHS))) +
  geom_point(position = position_jitter(
    height = 0.3, 
    width = 0.05), 
    size = 0.4, 
    color = "#0072B2", 
    alpha = 1/2) +
  geom_hpline(data = sampler(25, 
                             group = RACE), 
              height = 0.6, 
              color = "#D55E00") +
  theme_bw() + 
  transition_states(.draw, 1, 3)
```
