---
title: "3 Quantile–quantile plots"
subtitle: In-Class Exercise for Week 4
author: "KB"
editor: visual
date: 2023-05-06
execute:
  freeze: auto
  warning: false
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    code-line-numbers: true
---

## 1.Load the required packages

```{r}
pacman::p_load(rstatix,gt,patchwork,tidyverse,webshot2)
```

## 2.Load the data-set into R

```{r}
exam_data <- read_csv('data/Exam_data.csv', show_col_types = FALSE) 
```

## 3.Visualise Normal Distribution

Quantile--quantile (Q-Q) plots are a useful visualization when we want to determine to what extent the observed data points do or do not follow a given distribution. If the data is normally distributed, the points in a Q-Q plot will be on a straight diagonally line. Conversely, if the points deviate significantly from the straight diagonally line, then it's less likely that the data is normally distributed.

::: panel-tabset
## The Plot

```{r}
#| code-fold: false
#| echo: false
ggplot(exam_data, 
       aes(sample=ENGLISH))+
  stat_qq()+
  stat_qq_line()
```

## The Codes

```{r}
#| code-fold: false
#| eval: false
ggplot(exam_data, 
       aes(sample=ENGLISH))+
  stat_qq()+
  stat_qq_line()
```
:::

::: callout-note
## Note

We can see that the points deviate significantly from the straight diagonal line. This is a clear indication that the set of data is not normally distributed.
:::

## 4.Combining statistical graph and analysis table

We will need to install webshot2

```{r}
qq <-ggplot(exam_data, 
       aes(sample=ENGLISH))+
  stat_qq()+
  stat_qq_line()

sw_t <- exam_data %>%
  shapiro_test(ENGLISH) %>%
  gt()

tmp <- tempfile(fileext = '.png')
gtsave(sw_t,tmp)
table_png <- png::readPNG(tmp,
                          native = TRUE)

qq + table_png
```
